
\documentclass[12pt]{article}

\usepackage[left=2cm, right=2cm, top=2cm, bottom=2cm]{geometry}

\begin{document}

\pagenumbering{gobble}

\subsubsection*{Revisions to the paper ``Staged Compilation With Two-Level Type Theory''}
\vspace{1em}

\begin{itemize}
\item
  I added an appendix as a supplement, which includes a complete listing of 2LTT
  rules and equations in a more traditional derivation notation. The appendix
  also includes a more compact specification of the same thing using a
  second-order algebraic signature.
\item
  I extended the beginning of Section 3 with a motivation of algebraic models and quotients, and      a description of extracting algorithms via a setoid interpretation.
\item
  I added a bit more explanation to "Models and Syntax of 2LTT" in Section 3.3,
  trying to explain syntactic aspects.
\item
  I added Section 3.3.2 as a short discussion of elaborating surface syntax to formal syntax,
  as a justification for the usage of surface syntax.
\item
  I added Section 3.3.3 for a sketch of interpreting the syntax of 2LTT in setoids.
\item I rephrased the correctness of staging, now I talk about soundness,
  stability and "strictness" in Definition 4.2. Strictness means that staging
  strictly preserves all object-level type and term formers, where "strictly"
  refers to equality in underlying non-quotiented types and terms. This captures
  that staging should not compute object-level redexes. As far as I see, it is
  easy to show strictness of staging so I added it as Theorem 4.16.
\item
  I reworded Section 4.4 to be hopefully clearer and closer to what I actually
  did in the demo implementation.
\end{itemize}


%% Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

%% Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

%% Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\end{document}
